\chapter{Introduction}
\label{ch:introduction}

Corporate bankruptcy is a rare but consequential event. When a firm fails, the
losses cascade well beyond its own balance sheet: creditors write down loans,
suppliers lose receivables, employees lose income, and the disturbance can
propagate through the wider financial system. The ability to anticipate such
failures from publicly observable financial ratios is therefore of direct
economic value, and it has motivated a long line of quantitative research on
bankruptcy prediction. Since the seminal discriminant-analysis models of the
late 1960s, the field has moved steadily towards machine learning, which can
accommodate the high dimensionality and the non-linear interactions that
characterise firm-level financial data \parencite{liang2016}.

A recurring obstacle in this literature is class imbalance. In any realistic
sample, healthy firms vastly outnumber bankrupt ones; in the Taiwanese data used
throughout this thesis, only $3.23\%$ of the observations correspond to
bankruptcies. A classifier that simply predicts ``healthy'' for every firm would
already achieve an accuracy above $96\%$, yet it would be economically useless
because it never identifies the very cases that matter. Much of the recent
high-performance literature addresses this problem through synthetic
oversampling, most prominently the Synthetic Minority Over-sampling Technique
(SMOTE), and reports headline accuracies of $98$--$99\%$
\parencite{pham2025,suresh2025,sinap2025}. This thesis takes a deliberately
different stance. Synthetic oversampling fabricates bankruptcy observations that
never occurred in the market; the resulting models are then evaluated, in part,
on their ability to reproduce data that the researcher invented. We regard this
as scientifically questionable and instead train exclusively on real historical
observations, using undersampling of the majority class to restore balance.

\section{Research Question and Scope}
\label{sec:intro-rq}

The thesis pursues two connected questions. First, how well can bankruptcy be
predicted on the Taiwanese dataset using a clean, undersampling-only pipeline
that introduces no synthetic firms, and how does such a pipeline compare to the
undersampling benchmark of \textcite{wangliu2021}? Second, and more centrally,
which financial features do the resulting models rely on, and to what extent do
different feature-importance methods agree on that ranking? The analysis is
explicitly predictive rather than causal: it asks what the fitted models use,
not which economic mechanisms cause failure. Questions of causal
identification, hypothesis testing, and out-of-sample transfer to other markets
are left outside the scope and revisited only as limitations.

\section{Approach and Contribution}
\label{sec:intro-contribution}

Methodologically, the thesis builds directly on the undersampling framework of
\textcite{wangliu2021} --- the same Taiwanese dataset, the same repeated
stratified cross-validation, and the same emphasis on the $F_2$-score as the
primary metric, reflecting the asymmetric cost of missing a bankruptcy. Three
models of increasing complexity are compared: a logistic regression baseline, a
random forest, and a gradient-boosting classifier. The distinctive contribution
lies one layer deeper. Where \textcite{wangliu2021} stop at predictive
performance, this thesis adds an interpretability layer, comparing three
feature-importance perspectives --- native importance, permutation importance,
and Shapley additive explanations (SHAP) --- and analysing where and why their
rankings diverge. The central empirical finding is that this divergence is
systematic rather than incidental, and that it can be traced to well-understood
properties of the methods themselves.

The remainder of the thesis is organised as follows.
Chapter~\ref{ch:literature} situates the work within the bankruptcy-prediction
literature and carves out the interpretability gap it addresses.
Chapter~\ref{ch:dataset} describes the Taiwanese dataset.
Chapter~\ref{ch:methodology} develops the methodological backbone --- data
preprocessing, undersampling, the three prediction models, the cross-validation
design, the performance metrics, and hyperparameter optimisation.
Chapter~\ref{ch:featureimportance} introduces the three feature-importance
methods. Chapter~\ref{ch:results} reports the predictive results and the
importance rankings, with the divergence at its centre.
Chapter~\ref{ch:discussion} interprets these findings, and
Chapter~\ref{ch:conclusion} concludes.
