\chapter{Methodology}
\label{ch:methodology}

This chapter builds the methodological backbone of the thesis, while tackling the first forementioned problem of class imbalance. It evolves around the pipeline and tools used, which are grounded in the relevant scientific 
theory to build the prediction model. The Multicollinearity is addressed in the following chapter on Feature Importance. While others handle correlation through feature selection already at this stage 
(\textcite{pham2025}, \textcite{chaudhary2023}), this thesis takes a different approach by working with all the data to " ... preserve the integrity" (\textcite{wangliu2021})
and address the redundancy from high correlation in the main contribution of this thesis in Chapter~\ref{ch:featureimportance}.
The guiding idea is a pipeline that is reproducible at every step, follows established best practices, and trains only on real, historically observed firms rather than synthetic samples. 
% A fixed random seed keeps every result reproducible. Scaling and resampling happen strictly inside the cross-validation folds, so no information leaks from the test data into training.
The pipeline rests on three core components: data preprocessing, undersampling and the prediction models, which are all wrapped inside a cross-validation loop that, through an embedded hyperparameter search, 
gives us different performance metrics.


\section{Data Preprocessing}
\label{sec:preprocessing}

The only transformation applied to the continuous features is min--max
normalisation, which rescales each feature onto the unit interval $[0,1]$.
Writing $F$ for the set of values a given feature takes across all firms, every
value $x \in F$ is mapped to
\begin{equation}
  x' \;=\; \frac{x - \min(F)}{\max(F) - \min(F)}, \qquad \forall\, x \in F,
  \label{eq:minmax}
\end{equation}
where $\min(F)$ and $\max(F)$ are the smallest and largest values of that
feature. We follow the normalisation used by the dataset's originators
\parencite[562--563]{liang2016} and by \textcite[5]{wangliu2021}, and implement
it with scikit-learn's \texttt{MinMaxScaler}. Two features, the
\emph{Liability-Assets Flag} and the \emph{Net Income Flag}, are already binary
$0/1$ indicators rather than continuous ratios and are passed through unchanged.

The choice of scaling follows from the components we use rather than from the
data alone. The two tree ensembles are invariant to monotone transformations of
the features: their splits depend only on the rank order of the values, which a
transformation such as $\log(x)$ leaves unchanged. Two other components, however,
are scale-sensitive. Logistic regression is not scale-invariant --- under an
$\ell_1$ or $\ell_2$ penalty the scale of a feature affects how strongly its
coefficient is shrunk --- so a common scale is needed for the penalty to treat
all coefficients comparably; and the \emph{ClusterCentroids} undersampler
(Section~\ref{sec:undersampling}) relies on Euclidean distances, which are only
meaningful once features share a scale. Min--max normalisation supplies that
scale. This is where we depart from \textcite{wangliu2021}, who additionally
apply a logarithmic transformation to reduce skewness: that step suits their
distribution-sensitive naive-Bayes classifier, whereas our rank-based ensembles
and penalised linear model gain nothing from it, so we keep preprocessing to the
single min--max step. The scaler's minimum and maximum are estimated on each
fold's training portion only, never on the full dataset, so that no test
information leaks into training; the fold mechanics follow in
Section~\ref{sec:cv}.


% The only preprocessing applied to the continuous features is min--max
% normalisation, which rescales each feature to the unit interval. For a feature
% $x$ the transformed value is
% \begin{equation}
%   x_{\text{scaled}} = \frac{x - x_{\min}}{x_{\max} - x_{\min}},
%   \label{eq:minmax}
% \end{equation}
% where $x_{\min}$ and $x_{\max}$ are the minimum and maximum of that feature.
% Normalisation matters because the logistic regression is penalised and the
% undersampler relies on Euclidean distances; without a common scale, features
% with large numerical ranges would dominate both. The two categorical indicators,
% the \emph{Liability-Assets Flag} and the \emph{Net Income Flag}, are already
% binary and are passed through unchanged.

% Crucially, the parameters $x_{\min}$ and $x_{\max}$ in Equation~\eqref{eq:minmax}
% are estimated only on the training portion of each cross-validation fold and are
% then applied to the held-out portion. Fitting the scaler on the full sample
% would let information from the test fold seep into training and inflate the
% scores. As Section~\ref{sec:cv} details, this is enforced structurally by placing
% the scaler inside the cross-validation pipeline rather than applying it once to
% the whole dataset.

\section{Undersampling}
\label{sec:undersampling}

The imbalance documented in Chapter~\ref{ch:dataset} must be addressed before a
classifier can learn to recognise bankruptcies. Two routes are available.
Oversampling generates additional minority observations, whereas undersampling
removes majority observations until a more even ratio is reached. This thesis
uses undersampling exclusively, and this is a deliberate scientific commitment
rather than a technical convenience. Synthetic oversampling methods such as SMOTE
create bankruptcy observations that never occurred; a model trained and partly
evaluated on such data is asked to reproduce events the researcher invented. By
discarding majority observations instead, undersampling keeps every training
example a real, historically observed firm. The price is a smaller training set,
which the design below mitigates by retaining representative rather than random
majority firms.

\subsection{ClusterCentroids}
\label{sec:clustercentroids}

Concretely, the majority class is condensed using the \emph{ClusterCentroids}
procedure. The method runs $K$-means over the healthy firms, forming exactly as
many clusters as firms are to be retained. Each cluster contributes one
representative to the reduced majority set. With hard voting
(\texttt{voting=\textquotesingle hard\textquotesingle}), the representative is the
\emph{real} firm lying closest to the cluster centroid, not the centroid itself.
This distinction is essential: the default behaviour would return the arithmetic
centroids, which are synthetic averages of several firms and would silently
reintroduce exactly the kind of fabricated observation the design seeks to avoid.
Hard voting guarantees that the condensed majority set consists entirely of
genuine firms that are, by construction, spread across the feature space so as to
represent the healthy population well.

The degree of rebalancing is fixed by a sampling strategy of $0.3$, chosen for
comparability with \textcite{wangliu2021} rather than tuned. Under a stratified
ten-fold split, each training fold contains $198$ bankrupt firms; a ratio of
$0.3$ therefore retains $198 / 0.3 = 660$ healthy representatives, condensed from
the $5{,}939$ healthy firms available in the fold. Table~\ref{tab:fold}
summarises the resulting composition of a single training fold. The held-out test
fold is left untouched by the sampler --- it retains the natural class ratio ---
so that performance is always measured on a realistically imbalanced sample.

\begin{table}[htbp]
  \centering
  \caption{Class composition within one cross-validation fold after
    undersampling. The sampler acts only on the training partition.}
  \label{tab:fold}
  \begin{tabular}{lrrr}
    \toprule
    & \textbf{Healthy} & \textbf{Bankrupt} & \textbf{Total} \\
    \midrule
    Training (before) & $5{,}939$ & $198$ & $6{,}137$ \\
    Training (after)  & $660$     & $198$ & $858$ \\
    Test (unchanged)  & $660$     & $22$  & $682$ \\
    \bottomrule
  \end{tabular}
  \floatfoot{Training-fold ratio after undersampling: $76.9\%$ healthy /
    $23.1\%$ bankrupt.}
\end{table}

\section{Prediction Models}
\label{sec:models}

Three classifiers are compared, deliberately ordered from simple to complex so
that the added flexibility of each step can be judged against the performance it
buys. The comparison is framed by the bias--variance trade-off: the logistic
regression is a low-variance, higher-bias linear model, while the two tree
ensembles reduce error from opposite directions.

\subsection{Logistic Regression}
\label{sec:lr}

Logistic regression models the probability of bankruptcy as a sigmoid transform
of a linear index,
\begin{equation}
  P(y = 1 \mid \mathbf{x}) = \sigma(\boldsymbol{\beta}^\top \mathbf{x})
  = \frac{1}{1 + \exp(-\boldsymbol{\beta}^\top \mathbf{x})},
  \label{eq:logit}
\end{equation}
so that each coefficient $\beta_j$ is the change in the log-odds of bankruptcy
per unit change in feature $j$. The model is transparent and its coefficients
carry an economic sign, which makes it the natural interpretable baseline. To
cope with the many correlated features, an $\ell_1$ or $\ell_2$ penalty is added
to the log-likelihood; the $\ell_1$ penalty in particular drives some
coefficients to zero and thus performs an implicit selection. Its main
limitation is that it can represent only a linear decision boundary in the
feature space.

\subsection{Random Forest}
\label{sec:rf}

A random forest is an ensemble of decision trees whose predictions are averaged.
Each tree is grown on a bootstrap sample of the training data and, at every
split, is restricted to a random subset of features. Both devices decorrelate
the individual trees, so that averaging them cancels a large part of the variance
that a single deep tree would exhibit. The random forest therefore attacks the
error from the \emph{variance} side: individual trees are allowed to be flexible
and low-bias, and the ensemble tames their instability. Unlike the logistic
regression, trees capture threshold effects and interactions between features
without these having to be specified in advance.

\subsection{Gradient Boosting}
\label{sec:gb}

Gradient boosting also builds an ensemble of trees, but sequentially rather than
in parallel. Each new tree is fitted to the residual errors of the current
ensemble, so that the model gradually corrects its own mistakes. The base
learners are deliberately shallow --- in the extreme, single-split stumps --- so
that each contributes only a small refinement. Boosting thus attacks the error
from the \emph{bias} side: it combines many weak, high-bias learners into a
low-bias whole, with the learning rate and the number of trees controlling how
aggressively it does so.

\subsection{Why These Three}
\label{sec:model-choice}

Together the three models span a coherent methodological range: an interpretable
linear baseline, a variance-reducing bagging ensemble, and a bias-reducing
boosting ensemble. Alternatives that were considered but not pursued ---
support-vector machines, $k$-nearest neighbours, and neural networks --- either
duplicated the linear boundary of the logistic regression or added flexibility
without a commensurate gain on this dataset, at the cost of interpretability.
Restricting the comparison to these three keeps the interpretability analysis of
Chapter~\ref{ch:featureimportance} tractable, since each additional model
multiplies the number of importance rankings to be reconciled.

\section{Cross-Validation}
\label{sec:cv}

Performance is estimated by repeated stratified $k$-fold cross-validation, using
ten folds repeated three times, for a total of $30$ train/test evaluations per
model. Stratification preserves the $3.23\%$ bankruptcy rate in every fold, which
is indispensable at this level of imbalance, and the three repetitions average
out the sensitivity of the estimate to any particular partition.

The essential design decision concerns the order of operations. Both the min--max
scaler of Section~\ref{sec:preprocessing} and the undersampler of
Section~\ref{sec:undersampling} are fitted \emph{inside} each training fold and
never see the test fold. This is implemented through an \texttt{imbalanced-learn}
pipeline that chains scaler, sampler, and classifier; when the pipeline is
evaluated on a fold, the sampler is applied to the training partition only and is
automatically skipped at test time, while the scaler transforms the test
partition using parameters learned on the training partition. Chaining the steps
in this way makes leakage structurally impossible rather than merely a matter of
discipline. The hyperparameter search of Section~\ref{sec:hyperparam} is nested
within this same loop, so that every candidate configuration is subjected to the
full $30$-fit evaluation.

\section{Performance Metrics}
\label{sec:metrics}

All metrics derive from the confusion matrix, which cross-tabulates predicted
against actual classes into true positives (TP), false positives (FP), false
negatives (FN), and true negatives (TN), where the positive class is bankruptcy.
Table~\ref{tab:confusion-schema} fixes the notation.

\begin{table}[htbp]
  \centering
  \caption{Confusion-matrix notation, with bankruptcy as the positive class.}
  \label{tab:confusion-schema}
  \begin{tabular}{llcc}
    \toprule
    & & \multicolumn{2}{c}{\textbf{Predicted}} \\
    \cmidrule(lr){3-4}
    & & Bankrupt & Healthy \\
    \midrule
    \multirow{2}{*}{\textbf{Actual}}
      & Bankrupt & TP & FN \\
      & Healthy  & FP & TN \\
    \bottomrule
  \end{tabular}
\end{table}

From these counts, precision and recall are defined as
\begin{equation}
  \text{Precision} = \frac{\text{TP}}{\text{TP} + \text{FP}}, \qquad
  \text{Recall} = \frac{\text{TP}}{\text{TP} + \text{FN}},
  \label{eq:precrec}
\end{equation}
where precision is the share of predicted bankruptcies that are genuine and
recall is the share of genuine bankruptcies that are detected. The two are
combined by the $F_\beta$-score, a weighted harmonic mean,
\begin{equation}
  F_\beta = (1 + \beta^2)\,
  \frac{\text{Precision} \cdot \text{Recall}}
       {\beta^2 \cdot \text{Precision} + \text{Recall}},
  \label{eq:fbeta}
\end{equation}
which reduces to the familiar $F_1$ for $\beta = 1$. This thesis adopts
$\beta = 2$ as the primary metric. Setting $\beta = 2$ weights recall four times
as heavily as precision in Equation~\eqref{eq:fbeta}, reflecting the economic
asymmetry of the problem: failing to flag a firm that subsequently goes bankrupt
(a false negative) is far more costly than a false alarm that triggers closer
inspection of a healthy firm. The same choice is made by \textcite{wangliu2021},
which preserves comparability.

Accuracy and the area under the ROC curve are reported as secondary figures but
are never used for model selection. At a $96.8/3.2$ split, accuracy is dominated
by the majority class and rewards the trivial ``always healthy'' classifier,
while the ROC curve, which weights both classes symmetrically, likewise paints an
over-optimistic picture under strong imbalance. The $F_2$-score, focused on the
minority class, is the metric on which all decisions are based.

\section{Hyperparameter Optimisation}
\label{sec:hyperparam}

For each model, hyperparameters are selected by an exhaustive grid search nested
inside the cross-validation loop: every configuration is scored by its
fold-averaged $F_2$, and the best configuration is retained.
Table~\ref{tab:grids} lists the search spaces and the sizes of the resulting
grids.

\begin{table}[htbp]
  \centering
  \caption{Hyperparameter search spaces per model.}
  \label{tab:grids}
  \begin{tabular}{lll}
    \toprule
    \textbf{Model} & \textbf{Hyperparameters searched} & \textbf{Grid} \\
    \midrule
    Logistic Regression & penalty ($\ell_1$/$\ell_2$), $C$ & $12$ \\
    Random Forest       & \texttt{n\_estimators}, \texttt{max\_depth},
                          \texttt{max\_features}, \texttt{min\_samples\_leaf} & $36$ \\
    Gradient Boosting   & \texttt{learning\_rate}, \texttt{n\_estimators},
                          \texttt{max\_depth}, \texttt{subsample},
                          \texttt{min\_samples\_leaf} & $12$ \\
    \bottomrule
  \end{tabular}
  \floatfoot{Each configuration is evaluated over the full $30$-fit
    cross-validation.}
\end{table}

Two points deserve emphasis. First, the same nominal hyperparameter can mean
different things across models: a \texttt{max\_depth} of $10$ in the random
forest and of $2$ in gradient boosting are not comparable, because the forest
relies on deep, decorrelated trees whereas boosting relies on many shallow ones.
Depth values should therefore be read within, not across, models. Second, to
guard against overfitting, training and validation $F_2$ were traced as the key
complexity parameters were swept; the resulting complexity curves are collected
in Appendix~\ref{ch:appendix} and confirm that the selected configurations sit
at or before the onset of overfitting. The selected values and their scores are
reported in Chapter~\ref{ch:results}.
