\chapter{Feature Importance}
\label{ch:featureimportance}

Having established how the models are trained and evaluated, this chapter turns to
the interpretive question of the thesis, which financial features the fitted models
actually rely on. Three importance methods are applied to each of the three models:
native importance, permutation importance and SHAP. They are introduced here as
methods, while their numerical results are reported in Chapter~\ref{ch:results}.
These rankings describe what the models rely on to separate bankrupt from healthy
firms, not the mechanisms that cause insolvency and are therefore to be read as
associative rather than causal.

The second problem raised in Chapter~\ref{ch:dataset} returns here, because it
determines how the following rankings can be read. The redundancy among the $95$
ratios is structural, as all of them are derived from the same balance sheet and
income statement. \emph{Debt ratio \%} and \emph{Net worth/Assets} sum to exactly one
for every firm in the sample and are therefore the same quantity written twice,
while ROA(A), ROA(B) and ROA(C) differ only in the treatment of interest and
depreciation and correlate between $0.94$ and $0.99$. Across 
the whole feature set the correlation is nevertheless low, with a mean
absolute Pearson coefficient of $0.08$ and a median of $0.02$, which is consistent
with \textcite[5]{wangliu2021} and \textcite[7]{ansahnarh2024} reporting no substantial
multicollinearity on this data. Both observations hold at once because correlation
is measured pairwise. Of the $93$ continuous features, $31$ have at least one partner
above $0.9$, but each of them is redundant with a median of two partners out of $92$
and unrelated to the rest. These $28$ pairs are $0.65\%$ of all $4{,}278$ pairs, so
their near-perfect values disappear into an average dominated by the remaining ones.
A third of the features has a twin, and the global statistic still shows almost
nothing.

These features are kept rather than screened out, following \textcite{wangliu2021} in
preserving the integrity of the data. The redundancy therefore reaches the
interpretation stage intact and the three methods that follow each handle it
differently. That is what makes the comparison informative and it is why the
rankings in Chapter~\ref{ch:results} have to be read against one another rather than
in isolation.

\section{Native Importance}
\label{sec:fi-native}

Native importance reads the ranking directly off the fitted model and therefore
differs by model family. For the logistic regression it is the vector of coefficients from
Equation~\eqref{eq:lasso}. They carry a sign, and since the $\ell_1$ penalty has
already dropped the redundant twin from each correlated group, each surviving
coefficient stands on its own. This is the one model that reaches the interpretation stage with
the redundancy already resolved.

For the two tree ensembles it is the Gini impurity importance of
Section~\ref{sec:rf}, which is unsigned and carries two distortions. 
\textcite{strobl2007} show that impurity importance leans towards features that take
many different values, simply because they offer more points at which to split. With
$93$ continuous ratios, that bias applies here directly. 
Redundancy adds a second one. When two ratios are near-identical, a split can use
only one of them at a time, meaning across the forest the credit is shared between the
twins. The importance of what they jointly carry is split in two and each ends up
looking less important than the pair really is. Neither distortion shows up in the
ranking on its own, which is why the Gini importance is not read in isolation.

\section{Permutation Importance}
\label{sec:fi-permutation}

Permutation importance is model-agnostic and based on performance. After the models
are fitted, it asks how much the performance drops if we ``disregard'' one feature.
For one feature at a time, its values are shuffled across the firms, so that the
feature keeps its overall distribution but no longer lines up with which firms went
bankrupt. Everything else is left untouched and how far the $F_2$-score then drops
is recorded.

A feature the model genuinely relies on causes a large drop when shuffled, while a
feature it hardly uses would not move the score as much. Averaging this over ten repetitions then
makes sure this does not come down to a single lucky or unlucky shuffle,
following the same logic as cross-validation. And because the measure is defined
purely through the loss of the same $F_2$ the models were selected on, it applies
agnostically to all three and gives a common anchor for comparing them.

Its weakness is the redundancy from before. When a feature has a near-identical
twin, shuffling one of them changes little, because the twin still holds most of the
same information. Even a feature the model relies on can then look
unimportant. This is the counterpart to the impurity case where redundancy splits
the credit between twins, here it hides it. It is not a flaw in the method but 
rather a property of permutation under dependence and the reason why the three 
methods are compared against eachother in
Chapter~\ref{ch:results}.

\section{SHAP}
\label{sec:fi-shap}

SHAP (SHapley Additive exPlanations) bases feature importance on game theory and
differs from the previous two methods in the sense that it comes with axiomatic
guarantees \parencite{lundberg2017}. The idea is to rebuild a single prediction from
simple, additive pieces with a starting value plus one contribution for each feature.
\textcite{lundberg2017} write this as
\begin{equation}
  g(z') = \phi_0 + \sum_{i=1}^{95} \phi_i z_i',
  \label{eq:shap-additive}
\end{equation}

Here $g(z')$ is an explanation model, a replica that rebuilds the model's true
prediction by adding different pieces together. The first is the base value $\phi_0$,
an average prediction across all firms, before the model knows anything about the
specific firm. The others are contributions $\phi_i$ of the individual features,
added on top of that base. Each $\phi_i$ is the so-called Shapley value of feature $i$, through which we can see
just how important each feature is by its actual numeric contribution to the model's
prediction. Under the axioms behind SHAP these pieces return the true
prediction of the model exactly, which is the additive guarantee the method is named
for. The switch $z_i' \in \{0,1\}$ only states whether
feature $i$ is included ($1$) or left out ($0$).

The question is then how to find each contribution $\phi_i$, the Shapley value of
feature $i$. It comes from cooperative game theory as the fair way to divide the
``payoff'' of a prediction among the features, treating them as players that
cooperate. A feature's contribution is how much it changes the prediction when it
joins a group of the other features, or coalition, averaged over every possible
group. Writing $F$ for the full set of features, each term in the sum measures that
change for one group $S$ and the sum runs over every group, weighted by the
fraction in front:
\begin{equation}
  \phi_i = \sum_{S \subseteq F \setminus \{i\}}
  \frac{|S|!\,\bigl(|F| - |S| - 1\bigr)!}{|F|!}
  \left[\, f_{S \cup \{i\}}\!\left(x_{S \cup \{i\}}\right) - f_S\!\left(x_S\right) \right].
  \label{eq:shap-shapley}
\end{equation}
What is in the square bracket is that mentioned change. $f_{S \cup \{i\}}$ is the 
prediction from the
group $S$ together with feature $i$, read on their feature values
$x_{S \cup \{i\}}$. $f_S$ is the same prediction without $i$, so the difference
is what feature $i$ adds to the group. The fraction in front is a weight. It follows 
from the features joining the coalition
one at a time in a random order, where feature $i$'s effect depends on which features 
are already in the coalition $S$ when
it enters. The factorials just count how
many of these line-ups put exactly the group $S$ before $i$: $|S|!$ arranges the
features inside $S$, $(|F| - |S| - 1)!$ arranges the ones that come after $i$ and
$|F|!$ is the number of line-ups in total, so the fraction is how common that
arrangement is. With it, $\phi_i$ becomes feature $i$'s average contribution over
every order in which it could join, which is what makes the split fair. It is also
what makes it impossible to compute directly, because there is one group for every
combination of the other features, $2^{|F|-1}$ of them, so with $95$ features this is
$2^{94}$ groups per feature, far beyond anything that could be computed.

The way around this is to use the structure of each model to reach the same values
without ever calculating those $2^{94}$ groups. Both come from the \texttt{shap}
library \parencite{lundberg2017}. For the logistic regression, 
\texttt{LinearExplainer}\footnote{\url{https://shap.readthedocs.io/en/latest/generated/shap.LinearExplainer.html}}
reads the Shapley values directly from the coefficients. For the two tree ensembles,
\texttt{TreeExplainer}\footnote{\url{https://shap.readthedocs.io/en/latest/generated/shap.TreeExplainer.html}} 
walks through the trees and recovers the exact values in a
tiny fraction of the time. In both cases the values are computed for the 
bankruptcy class across all firms and are exact Shapley values rather than 
approximations.

Two properties make SHAP the interpretive center of this thesis. First, it returns
signed contributions for all three models, where the logistic coefficients are signed
but only linear and the impurity importance has no sign at all, SHAP puts every model
on the same signed scale and makes them directly comparable. Second, because
it works with whole groups of features rather than changing one at a time, it does not
collapse under the redundancy that impacts permutation importance. Although it is not fully
immune to that redundancy, with near-identical twins it shares their joint
contribution between them, so a redundant pair is spread across two smaller entries.
But unlike impurity, which splits the credit arbitrarily, or permutation, which
hides it, SHAP keeps both the total and the sign, so the pair stays readable. This is
why SHAP gives the most trustworthy reading, but is still compared against the other
two rather than trusted alone in Chapter~\ref{ch:results}. It is the pattern across 
them, not any single method, that carries the insight.