\chapter{Discussion}
\label{ch:discussion}

The results invite interpretation on two levels: what the divergence between the
importance methods tells us about the methods and the data, and what the rankings
tell us --- cautiously --- about bankruptcy itself. This chapter takes each in
turn and then revisits the limitations that bound both readings.

\section{Reading the Divergence}
\label{sec:disc-divergence}

The most robust finding of the thesis is that the three importance methods
disagree in a structured, explicable way, and that the disagreement is
diagnostic rather than merely inconvenient. The elevation of \emph{Persistent
EPS} to first place by the native Gini importance of both ensembles is best read
as a method artefact: it is the impurity bias of \textcite{strobl2007} operating
exactly as theory predicts on a fine-grained continuous feature. That SHAP demotes
this feature and promotes \emph{Borrowing dependency} for gradient boosting is
correspondingly reassuring, because it is the method with the stronger
theoretical guarantees that yields the more economically coherent ordering. The
lesson generalises beyond this dataset: reporting a single native importance
ranking, as much of the applied literature does, risks mistaking a property of
the estimator for a property of the world.

The collapse of permutation importance on the ensembles admits a symmetric
reading. Taken at face value its near-zero values would suggest that the model
relies on almost nothing, which cannot be reconciled with the model's evident
predictive success. The resolution is correlation masking: with $95$ collinear
ratios, no single feature is indispensable because a correlated substitute
always remains. This makes permutation importance a poor ranking device on this
data but an excellent \emph{diagnostic} of its redundancy. Read together, then,
the two performance-based methods are complementary --- SHAP for the ranking,
permutation for the warning about collinearity --- and it is the comparison, not
any single method, that delivers the insight. This is the sense in which the
thesis widens rather than opens the black box: it does not claim a single correct
importance list but shows where each method can and cannot be trusted.

\section{An Economic Reading, Under Caveat}
\label{sec:disc-economic}

If the SHAP rankings are read economically --- and only associatively --- they
are broadly sensible. Leverage and debt-servicing capacity dominate: the debt
ratio for the logistic regression, and borrowing dependency, the degree of
financial leverage, and interest-coverage measures for gradient boosting. That
firms carrying more debt relative to their capacity to service it are more likely
to fail is consistent with basic corporate-finance intuition, and it is
encouraging that a pipeline trained purely for prediction recovers this structure
without being told to. The appearance of profitability and cash-position measures
alongside leverage rounds out a coherent, if unsurprising, financial picture.
The emphasis must nonetheless remain on \emph{association}: these features are
what the models lean on to separate the classes, not established causes of
insolvency, and the analysis contains no design capable of supporting the
stronger claim.

\section{Comparison with Wang and Liu and Limitations}
\label{sec:disc-limits}

Against the reference benchmark, the results are competitive but must be framed
honestly. The best configuration here, gradient boosting with cluster-centroid
undersampling at $F_2 = 0.550$, exceeds the best figure reported by
\textcite{wangliu2021}, whose edited-nearest-neighbours and naive-Bayes
combination reached $F_2 = 0.423$. This is not a claim to have beaten them under
identical conditions: the two studies use related but distinct undersampling
algorithms, and the pairing of tree ensembles with cluster-centroid
undersampling is simply a combination they did not explore. The contribution is
therefore twofold --- a competitive clean-pipeline result, and the
interpretability layer that their work lacked entirely.

Several limitations temper the conclusions. The associative nature of the
importance analysis has been stressed repeatedly and is the most important of
them. Beyond it, the dataset offers no macroeconomic controls and covers a single
market and period, so the transferability of both the performance and the
rankings to other settings is untested. Undersampling, while protecting the
integrity of the training data, discards the majority of healthy firms and so
sacrifices information that a larger sample might exploit. And the sampling ratio
was fixed rather than tuned, for comparability; a fuller study would treat it as
a design parameter. None of these qualifications undermines the central result,
but each marks a boundary of what the present analysis can claim.
