\chapter{Conclusion}
\label{ch:conclusion}

This thesis set out to predict corporate bankruptcy on the Taiwanese dataset
using a pipeline that admits no synthetic data, and to open the resulting models
to interpretation. On the first question, a clean undersampling pipeline built on
cluster-centroid sampling and evaluated by the $F_2$-score achieves a best
cross-validated score of $0.550$ for gradient boosting, ahead of the random
forest and well ahead of the logistic-regression baseline, and competitive with
the undersampling benchmark of \textcite{wangliu2021} while training exclusively
on real firms. On the second, and more central question, comparing native,
permutation, and SHAP importance across the three models reveals a systematic
divergence: native impurity importance is distorted by the bias of
\textcite{strobl2007}, permutation importance collapses under the collinearity of
the $95$ ratios, and SHAP --- the only method that yields signed, comparable
attributions for every model --- supplies the most trustworthy ranking, in which
leverage and debt-servicing measures dominate.

The methodological contribution is the pairing of a scientifically honest
sampling design with an explicit, multi-method interpretability analysis. Rather
than reporting a single importance list, the thesis treats the disagreement
between methods as its object of study and shows that this disagreement is
predictable from the known properties of each method and from the structure of
the data. For the practitioner, the message carries a modest policy relevance:
an early-warning model can be built on real data alone, without fabricating
distress events, and its outputs can be interrogated for the financial signals
that drive them --- provided those signals are read as predictive associations
and not as causes on which intervention could naively be based.

Two extensions suggest themselves. Replicating the analysis on a second dataset
from another market would test whether both the performance and the importance
rankings transfer, or whether they are specific to the Taiwanese sample. And a
deeper economic treatment --- ideally one that moves from association towards a
design capable of causal statements --- would let the leverage-centred picture
recovered here be assessed as more than a description of what the models happen to
use. Both lie beyond the present scope but follow naturally from its results.
